Other metrics

### 2021-02-24 analysis/rev_prototyping/additional_metrics_results.Rmd
some updates to di
run

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(drake)
library(dplyr)
library(ggplot2)
library(grid)
theme_set(theme_bw())

@@

all_di <- all_di %>%
  filter(s0 > 1,
         s0 != n0,
         n0 != s0 + 1)

@@

@@

ggplot(filter(all_di), aes(mean_po_comparison_percentile)) +
  geom_histogram() +
  facet_wrap(vars(dat, singletons), scales = "free_y")

all_di %>% group_by(singletons, dat) %>% summarize(po_sig = mean(mean_po_comparison_percentile > 95))

ggplot(all_di, aes(mean_po_comparison_95, mean_po_comparison, color = log(nparts))) +
  geom_point() +
  geom_line(aes(mean_po_comparison_95, mean_po_comparison_95))

all_di %>%
  filter(mean_po_comparison_percentile > 95) %>%
  group_by(singletons) %>%
  summarize(diff = mean(mean_po_comparison - mean_po_comparison_95))

ggplot(all_di, aes(nparts, mean_po_comparison_95_ratio_1t)) +
  geom_point() +
  scale_x_log10()

ggplot(all_di, aes(nparts, mean_po_comparison_95_ratio_1t)) +
  geom_point() +
  scale_x_log10() +
  xlim(0, 10000)

@@

@@

ggplot(all_di, aes(mean_po_comparison_sims, mean_po_comparison, color = mean_po_comparison_percentile > 95)) +
  geom_point(alpha = .3) +
  facet_wrap(vars(dat, singletons))

ggplot(all_di, aes(mean_po_comparison - mean_po_comparison_sims, color = mean_po_comparison_percentile > 95)) +
  geom_histogram(alpha = .3) +
  facet_wrap(vars(dat, singletons))

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.dim =c(6,6))
library(drake)
library(dplyr)
library(ggplot2)
library(grid)
theme_set(theme_bw())

@@

mcdb_new_di <-read.csv(here::here("mcdb_new_di.csv"))

all_di <- all_di %>%
  filter(s0 > 1,
         s0 != n0,
         n0 != s0 + 1,
         !singletons)

@@

Defined as the proportion of individuals allocated to species of different abundances. Most intuitive as a graph - here are two hypothetical SADs for a community with 7 species and 71 individuals:

@@

## Set up the cache and config
db <- DBI::dbConnect(RSQLite::SQLite(), here::here("analysis", "drake", "drake-cache-net.sqlite"))
cache <- storr::storr_dbi("datatable", "keystable", db)
cache$del(key = "lock", namespace = "session")
knitr::opts_chunk$set(fig.width=3, fig.height=2.5, warning = FALSE, message = FALSE)

example_fs <- readd(fs_s_7_n_71, cache = cache) %>%
  filter(sim %in% c(1, 2)) %>%
  mutate(sim = paste0("sim_", sim))

example_fs_wide <- example_fs %>%
  select(sim, abund, rank) %>%
  tidyr::pivot_wider(id_cols = rank, names_from = sim, values_from = abund)

ggplot(example_fs_wide, aes(rank, sim_1)) +
  geom_point() +
  geom_line() +
  geom_point(aes(rank, sim_2), color = "blue") +
  geom_line(aes(rank, sim_2), color = "blue") +
  geom_ribbon(aes(rank, ymin = sim_1, ymax = sim_2), fill = "grey", alpha = .2) +
  ylab("abund")
```

We take the grey area - the area of difference between the two SADs - and divide it by 2 (because every individual allocated to a different species will count twice), and divide that total by the total number of individuals in the community. In principle this metric ranges from 0 to 1, with 0 being no individuals allocated differently and 1 being all individuals allocated differently, although note that neither 0 nor 1 can actually be achieved.

Here are those calculations:

```{r, echo = T}
(sum(abs(example_fs_wide$sim_1 - example_fs_wide$sim_2)) / 2) / sum(example_fs_wide$sim_1)

fs_mat <- select(example_fs_wide, sim_1, sim_2) %>% t()

scadsanalysis::proportion_off(fs_mat)

@@

This is a metric of dissimiliarty defined for *two* focal vectors. We want to know whether the observed SAD is more unlike the elements of its feasible set than the elements of the feasible set are unlike each other. To do this, we calculate the proportion off between the observed SAD and a large number of samples from the feasible set, and take the mean score of all these comparisons. This tells us how different, on average, the observed SAD is from samples. We then repeat this process many times, but instead of the observed SAD we select a random sample from the feasible set as focal sample, compare this focal sample to many other samples, and take the mean. The distribution of these scores can then be compared directly to the score from the observed sample.

```{r}
ggplot(filter(all_di, nparts > 20), aes(mean_po_comparison_percentile)) +
  geom_histogram() +
  facet_wrap(vars(dat, singletons), scales = "free_y")

all_di %>% filter(nparts >20) %>% group_by(singletons, dat) %>% summarize(high_proportion_off = mean(mean_po_comparison_percentile > 95))

Because the units for this metric do not depend on S and N, we can also estimate the effect size as the difference in the proportion off between the observed SAD compared to the feasible set and the proportion off for elements of the feasible set compared to each other.

Looking at those communities where the observed SAD is much more unlike the elements of the feasible set, than the elements of the FS are unlike each other, we can ask how much *more* dissimilar the SAD is than the 95th percentile of unlikeness from the feasible set:

@@

all_di %>%
  filter(mean_po_comparison_percentile > 95) %>%
  group_by(singletons, dat) %>%
  summarize(diff_from_95 = mean(mean_po_comparison - mean_po_comparison_95))

ggplot(filter(all_di, mean_po_comparison_percentile > 95), aes( mean_po_comparison - mean_po_comparison_95)) +
  geom_histogram() +
  facet_wrap(vars(dat), scales = "free_y")

@@

That is, e.g. for BBS, the significant deviation is that the observed-to-FS dissimilarity is on average .07 higher than the 95th percentile of FS-to-FS dissimilarity.

We can examine how a one-tailed 95% breadth index changes over the size of the feasible set:

ggplot(all_di, aes(nparts, mean_po_comparison_95_ratio_1t, color = dat)) +
  geom_point(alpha= .2) +
  scale_x_log10() +
  theme(legend.position = "top")
# 
# 
# ggplot(all_di, aes(nparts, mean_po_comparison_95_ratio_1t, color = dat)) +
#   geom_point(alpha= .8) +
#   scale_x_log10() +
#   xlim(c(0, 30000))+
#   theme(legend.position = "top")
# 
# 
# ggplot(all_di, aes(nparts, mean_po_comparison_95_ratio_1t, color = dat)) +
#   geom_point(alpha= .8) +
#   scale_x_log10() +
#   xlim(c(0, 2000))+
#   theme(legend.position = "top")

<!-- Or we can ask what the difference in dissimilarity scores is on average: -->

<!-- ```{r} -->
<!-- all_di %>% -->
<!--   group_by(singletons, dat) %>% -->
<!--   summarize(mean_actual = mean(mean_po_comparison), -->
<!--             mean_sims = mean(mean_po_comparison_sims), -->
<!--             mean_diff = mean(mean_po_comparison - mean_po_comparison_sims)) -->

<!-- ggplot(filter(all_di), aes(mean_po_comparison_sims, mean_po_comparison, color =  mean_po_comparison_percentile > 95)) + -->
<!--   geom_point() + -->
<!--   facet_wrap(vars(dat)) + -->
<!--   geom_abline(slope = 1, intercept = 0) + -->
<!--   theme(legend.position = "top") -->
<!-- #  -->
<!-- # po_long <- all_di %>% -->
<!-- #   select(dat, singletons, mean_po_comparison, mean_po_comparison_sims) %>% -->
<!-- #   tidyr::pivot_longer(cols = c("mean_po_comparison", "mean_po_comparison_sims"), names_to = "comp", values_to = "val")  %>% -->
<!-- #   mutate(source = ifelse(grepl("sims", comp), "fs_to_fs", "obs_to_fs")) -->
<!-- #  -->
<!-- # ggplot(po_long, aes(source, val)) + -->
<!-- #   geom_boxplot() + -->

<!-- ggplot(filter(all_di), aes( mean_po_comparison - mean_po_comparison_sims)) + -->
<!--   geom_histogram() + -->
<!--   facet_wrap(vars(dat), scales = "free_y") -->

<!-- ``` -->

<!-- ```{r} -->

<!-- ggplot(all_di, aes(mean_po_comparison_sims, mean_po_comparison, color = mean_po_comparison_percentile > 95)) + -->
<!--   geom_point(alpha = .3) + -->
<!--   facet_wrap(vars(dat, singletons)) -->

<!-- ggplot(all_di, aes(mean_po_comparison - mean_po_comparison_sims, color = mean_po_comparison_percentile > 95)) + -->
<!--   geom_histogram(alpha = .3) + -->
<!--   facet_wrap(vars(dat, singletons)) -->

<!-- ``` -->
